% =====================================================================
%  Market Regime Detection Using Autoencoders and PCA
%  Presentation Script (15 min) + Q&A Preparation
%
%  This document is a SPOKEN WALKTHROUGH OF THE REPORT (main.tex), not of
%  the code. Each block maps to a section of the paper, so the talk and the
%  written report tell exactly the same story: motivation -> data ->
%  method -> results -> trading test -> limitations -> future work.
%
%  Compile with pdflatex (run twice for the table of contents).
%  Only standard packages are used (works on Overleaf out of the box).
% =====================================================================
\documentclass[11pt]{article}

\usepackage[a4paper,margin=2.2cm]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{titlesec}
\usepackage{enumitem}
\usepackage{fancyhdr}
\usepackage{array}
\usepackage{booktabs}
\usepackage[most,skins,breakable]{tcolorbox}
\usepackage{hyperref}

% ---------- Colour palette -------------------------------------------
\definecolor{ink}{HTML}{1A1A2E}
\definecolor{soft}{HTML}{F4F6FB}
\definecolor{rule}{HTML}{D7DEEA}

% Person 1 — Data (blue)
\definecolor{p1}{HTML}{1565C0}
\definecolor{p1l}{HTML}{E3F0FC}
% Person 2 — Autoencoder (teal/green)
\definecolor{p2}{HTML}{00897B}
\definecolor{p2l}{HTML}{DEF3F0}
% Person 3 — PCA (purple)
\definecolor{p3}{HTML}{6A1B9A}
\definecolor{p3l}{HTML}{F1E6F8}
% Accents
\definecolor{amber}{HTML}{E08A00}
\definecolor{amberl}{HTML}{FBEFD6}
\definecolor{crimson}{HTML}{C62828}

% current accent (switched per section)
\colorlet{accent}{p1}
\colorlet{accentl}{p1l}

% ---------- Section styling ------------------------------------------
\titleformat{\section}
  {\sffamily\LARGE\bfseries\color{accent}}{}{0pt}
  {}[{\color{accent}\titlerule[2pt]}]
\titlespacing*{\section}{0pt}{18pt}{8pt}

\titleformat{\subsection}
  {\sffamily\large\bfseries\color{accent}}{}{0pt}{}
\titlespacing*{\subsection}{0pt}{12pt}{4pt}

\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt}
\renewcommand{\familydefault}{\sfdefault}

% ---------- Header / footer ------------------------------------------
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0.6pt}
\renewcommand{\headrule}{\hbox to\headwidth{\color{rule}\leaders\hrule height \headrulewidth\hfill}}
\fancyhead[L]{\footnotesize\color{ink}\textbf{Market Regime Detection} — AE \& PCA}
\fancyhead[R]{\footnotesize\color{accent}\leftmark}
\fancyfoot[C]{\footnotesize\color{ink}\thepage}

\hypersetup{colorlinks=true, linkcolor=accent, urlcolor=p1}

% ---------- Reusable boxes -------------------------------------------
% Timing cue box: \begin{cue}{label}{duration} ... \end{cue}
\newtcolorbox{cue}[2]{
  enhanced, breakable, sharp corners=downhill,
  colback=accentl, colframe=accent, boxrule=0pt,
  leftrule=4pt, top=5pt, bottom=5pt, left=9pt, right=9pt,
  before skip=8pt, after skip=8pt,
  overlay unbroken and first={
    \node[anchor=north east, font=\sffamily\bfseries\footnotesize,
          text=white, fill=accent, inner sep=3pt, rounded corners=2pt]
      at ([xshift=-6pt,yshift=-5pt]frame.north east) {#2};},
  title={\sffamily\bfseries\color{accent}#1},
  fonttitle=\sffamily\bfseries
}

% Slide-cue box: which figure/table from the report to show
\newtcolorbox{slide}{
  enhanced, breakable, sharp corners,
  colback=white, colframe=rule, boxrule=1pt,
  left=9pt, right=9pt, top=4pt, bottom=4pt,
  before skip=6pt, after skip=6pt
}

% Q&A box: \begin{qa}{question} answer \end{qa}
\newtcolorbox{qa}[1]{
  enhanced, breakable, sharp corners,
  colback=white, colframe=accent, boxrule=1pt,
  left=10pt, right=10pt, top=8pt, bottom=8pt,
  before skip=7pt, after skip=7pt,
  coltitle=white, colbacktitle=accent,
  title={\sffamily\bfseries Q.\hspace{4pt}#1}
}

% Key-numbers card
\newtcolorbox{numbers}[1]{
  enhanced, sharp corners,
  colback=soft, colframe=accent, boxrule=0pt, leftrule=4pt,
  left=10pt, right=10pt, top=6pt, bottom=6pt,
  title={\sffamily\bfseries\color{accent}#1}, fonttitle=\sffamily\bfseries
}

% Person banner
\newcommand{\personbanner}[3]{%
  \begin{tcolorbox}[enhanced, sharp corners, colback=accent, colframe=accent,
      left=14pt, right=14pt, top=10pt, bottom=10pt, before skip=4pt, after skip=10pt]
    {\sffamily\bfseries\LARGE\color{white} #1}\\[2pt]
    {\sffamily\color{white}\large #2}\hfill{\sffamily\bfseries\color{white} #3}
  \end{tcolorbox}}

\newcommand{\setp}[2]{\colorlet{accent}{#1}\colorlet{accentl}{#2}}

% Small helper: "Report ref" tag inside a slide cue
\newcommand{\reportref}[1]{{\small\color{ink}\textbf{Show:} #1}}

% =====================================================================
\begin{document}

% ---------------- Title page -----------------------------------------
\begin{titlepage}
\setp{p1}{p1l}
\centering
\vspace*{2cm}
{\sffamily\bfseries\Huge\color{ink} Market Regime Detection}\\[6pt]
{\sffamily\bfseries\Huge\color{p1} Using Autoencoders \& PCA}\\[18pt]
{\color{rule}\rule{0.7\textwidth}{1.5pt}}\\[18pt]
{\sffamily\Large\color{ink} Presentation Script \;\&\; Q\&A Preparation}\\[6pt]
{\sffamily\large\color{ink} A spoken walkthrough of the written report}\\[4pt]
{\sffamily\large\color{ink} 15-minute group talk \,$\cdot$\, three speakers \,$\cdot$\, 5 minutes each}\\[40pt]

\begin{tcolorbox}[enhanced, width=0.82\textwidth, halign=center,
   colback=soft, colframe=rule, boxrule=1pt, sharp corners]
\sffamily\large
\textcolor{p1}{\textbf{Person 1 — Harjot Singh}}\\ Motivation, Data \& Preprocessing\\ {\small (Report §I–III)}\\[8pt]
\textcolor{p2}{\textbf{Person 2 — Eleni Tsaousi}}\\ Autoencoder, Training \& Clustering\\ {\small (Report §IV; §V-A,B)}\\[8pt]
\textcolor{p3}{\textbf{Person 3 — Adene Dinoshi}}\\ PCA, Trading Strategy, Limitations \& Outlook\\ {\small (Report §V-C–F; §VI–VIII)}
\end{tcolorbox}

\vfill
{\sffamily\small\color{ink} Investments — Selected Quantitative Tools, University of Zurich \\ Supervisor: Matthias Feiler}
\end{titlepage}

% ---------------- Table of contents ----------------------------------
\setp{ink}{soft}
{\hypersetup{linkcolor=ink}
\renewcommand{\contentsname}{Contents}
\tableofcontents}
\vspace{6pt}

\begin{tcolorbox}[enhanced, sharp corners, colback=amberl, colframe=amber,
    boxrule=0pt, leftrule=4pt, left=10pt, right=10pt, top=6pt, bottom=6pt]
\sffamily\small
\textbf{\textcolor{amber}{How to use this document.}} This is a presentation of
\emph{the report}, section by section --- not a tour of the code. Part~I is the
word-for-word script, split into three colour-coded sections (one per speaker),
each with a timing cue, a ``key numbers'' card, and a grey \textbf{Show} cue
naming the exact figure or table from the report to put on screen. Part~II is the
Q\&A: master your own colour-coded section and skim the others. All figures and
tables referenced are those in the report
(\texttt{Market\_Regime\_Detection\_Using\_Autoencoders\_and\_PCA/}, mirrored in
\texttt{pca-regime-analysis/outputs/figures/}).
\end{tcolorbox}

\newpage

% #####################################################################
% PART I — SCRIPT
% #####################################################################
\setp{ink}{soft}
\begin{center}
{\sffamily\bfseries\Huge\color{ink} Part I \;\textemdash\; Presentation Script}
\end{center}
\vspace{4pt}

% =====================================================================
% PERSON 1
% =====================================================================
\setp{p1}{p1l}
\section{Person 1 — Motivation \& Data (Report §I–III)}
\personbanner{Harjot Singh}{Motivation, Data \& Preprocessing}{$\approx$ 5 min}

\begin{numbers}{Key numbers (from the report)}
\sffamily\small
Weekly data, \textbf{1988 → 2026} ($\approx$38 yrs) \;$\cdot$\;
\textbf{30 features} (8 log-return + 22 macro/valuation) \;$\cdot$\;
Split \textbf{70/15/15 by time} (train 1988–2014, val 2014–2020, test 2020–2026) \;$\cdot$\;
Window \textbf{26 weeks}, stride 1, flattened to \textbf{780} \;$\cdot$\;
Train tensor \textbf{1365 $\times$ 26 $\times$ 30}.
\end{numbers}

\begin{cue}{Opening \& roadmap}{40 sec}
Good morning. I'm Harjot, and on behalf of our group I'll walk you through our
report on \textbf{market regime detection}. Here's the plan for the next fifteen
minutes: I'll set up the \emph{problem} and the \emph{data} the report is built
on; Eleni will present the \emph{autoencoder} and how we turn it into regimes; and
Adene will cover the \emph{PCA benchmark}, the \emph{trading test} of whether
those regimes are worth anything, and our honest \emph{limitations and next
steps}. The whole talk follows the structure of the written report.
\end{cue}

\begin{cue}{The question \& objective}{60 sec}
The report opens with one question: can we detect distinct \textbf{market
regimes} --- recurring states like calm bull markets, inflationary periods, or
crises --- from financial and macroeconomic data alone, with \emph{no} manual
labelling? This matters because knowing the prevailing regime supports portfolio
allocation, risk management, and asset pricing.

Our objective, stated in Section~I, is to test whether an \textbf{autoencoder} can
learn a meaningful low-dimensional representation of market behaviour that
clusters into regimes --- and, crucially, to benchmark it against \textbf{PCA} at
identical dimensionality, so any difference is attributable to the \emph{method},
linear versus non-linear, not to the number of dimensions.
\end{cue}

\begin{slide}\reportref{Section~III ``Data and Preprocessing'' --- the dataset description}\end{slide}
\begin{cue}{The data}{50 sec}
Section~III describes the data. We use the course market dataset: a weekly US
series from 1988 to 2026 --- nearly 38 years and multiple business cycles. It has
\textbf{30 features} spanning two groups: eight price/index series (the equity
market, gold, oil, the dollar, Treasury and credit prices) and twenty-two macro
and valuation indicators (yield curve, CPI, unemployment, CAPE, dividend yield,
GDP). Together they capture both \emph{what} markets are doing and the economic
environment \emph{behind} it --- exactly what regime detection needs.
\end{cue}

\begin{cue}{Making the data stationary}{50 sec}
The first preprocessing problem is \textbf{non-stationarity}. The eight price
columns are raw levels that simply trend upward for decades; fed in directly, the
model would learn the \emph{trend}, not the \emph{regime}. Augmented
Dickey--Fuller tests put their $p$-values near 1.0 --- strongly non-stationary.
The fix is to convert each to \textbf{log returns}; afterwards all eight have
$p<0.0001$. The 22 macro features are already stationary rates and ratios, so we
keep them as they are.
\end{cue}

\begin{cue}{Splits \& anti-leakage}{50 sec}
The second principle, also in Section~III, is preventing \textbf{look-ahead
leakage}. We split \textbf{70/15/15 by time, never shuffled}: train is the oldest
data (1988–2014), then validation (2014–2020), and the test set is the most recent
period (2020–2026). A \texttt{StandardScaler} is fit on \emph{training data only}
and then applied to validation and test --- so the model never uses the future to
standardise the past. This forward-chaining design is what makes our
out-of-sample claims honest.
\end{cue}

\begin{cue}{Rolling windows \& the handoff}{50 sec}
Finally, the input format. A regime isn't a single week --- it's a stretch of
time --- so we slide a \textbf{26-week window} (about six months) across each split
with stride one, never crossing a split boundary. Each window flattens
$26 \times 30 = \textbf{780}$ values, and the training tensor is
$1365 \times 26 \times 30$. Twenty-six weeks is long enough to smooth single-week
noise yet short enough to catch a regime change within a quarter or two. I handed
Eleni a clean, leak-free, model-ready dataset.
\end{cue}

\begin{cue}{Hand over}{10 sec}
With that foundation in place, I'll hand over to Eleni, who turns these windows
into a learned representation and then into regimes.
\end{cue}

% =====================================================================
% PERSON 2
% =====================================================================
\setp{p2}{p2l}
\section{Person 2 — Autoencoder \& Clustering (Report §IV; §V-A,B)}
\personbanner{Eleni Tsaousi}{Method, Training, Reconstruction \& Clustering}{$\approx$ 5 min}

\begin{numbers}{Key numbers (from the report)}
\sffamily\small
Input \textbf{780} ($26\times30$) \;$\cdot$\;
Encoder \textbf{780→256→64→16}, symmetric decoder, ReLU \;$\cdot$\;
Latent \textbf{16-D} \;$\cdot$\;
Adam, weight decay \textbf{5e-4}, early stopping (patience 40), $\le$200 epochs \;$\cdot$\;
Recon MSE \textbf{train 0.30 / val 0.61 / test 1.20} \;$\cdot$\;
Ablation: \textbf{16-D} gives the lowest test MSE (\textbf{1.20}) \;$\cdot$\;
KMeans \textbf{K=4} (fit on train latent only).
\end{numbers}

\begin{cue}{Opening}{20 sec}
Thank you, Harjot. I'm Eleni. I'll present the methodology and the autoencoder
results --- Sections~IV and~V-A,B of the report --- which is the heart of the
pipeline: how we compress those windows and turn them into four regimes.
\end{cue}

\begin{slide}\reportref{Fig.~1 ``framework'' --- the four-stage pipeline diagram}\end{slide}
\begin{cue}{The framework}{40 sec}
The report frames the method as \textbf{four stages}, shown in Figure~1: compress
each window into a low-dimensional representation --- with \emph{both} an
autoencoder and PCA --- then cluster the representations with K-Means, then
interpret each cluster as a market regime. Everything is \textbf{unsupervised}:
regimes aren't labelled in advance, we let the structure emerge.
\end{cue}

\begin{slide}\reportref{Fig.~2 ``autoencoder\_architecture'' --- encoder/bottleneck/decoder}\end{slide}
\begin{cue}{Why an autoencoder, and its architecture}{60 sec}
An autoencoder compresses its input through a narrow \textbf{bottleneck} and
reconstructs it; the bottleneck forces it to keep only what's essential. Its
advantage over PCA is capturing \textbf{non-linear} structure --- and markets are
non-linear: a rise in yields means something different when inflation is high
versus low. Figure~2 shows the architecture: the 780-value window is compressed
$256 \to 64 \to \textbf{16}$ with ReLUs, then a symmetric decoder rebuilds it to
780. We train on \textbf{mean squared reconstruction error}. The report justifies
16 latent dimensions on two grounds: the features are highly correlated, so the
effective dimensionality is far below 30; and a much narrower bottleneck causes
\emph{mode collapse}.
\end{cue}

\begin{slide}\reportref{Table~II ``reconstruction loss by split''}\end{slide}
\begin{cue}{Training \& reconstruction results}{50 sec}
We train with Adam and weight decay for regularisation, and --- importantly ---
\textbf{early stopping} on validation loss, keeping the best checkpoint. Section~V-A
reports the reconstruction loss: about \textbf{0.30} on train, \textbf{0.61} on
validation, \textbf{1.20} on test --- and roughly a 70\% improvement over a naive
mean-predicting baseline, confirming the model learns genuine structure. The much
higher \emph{test} loss is expected: 2020–2026 includes COVID and the post-pandemic
inflation-and-rate-hike cycle. The report reads this as evidence of a possible
\textbf{regime shift}, not a model failure --- a thread Adene will pick up.
\end{cue}

\begin{slide}\reportref{Table~III ``latent-dimension ablation''}\end{slide}
\begin{cue}{Ablation: why 16 dimensions}{50 sec}
Section~V-B stress-tests that choice. We retrain at latent widths 2, 3, 8, 16 and
32. Two findings: \textbf{16 dimensions} gives the \emph{lowest} test reconstruction
loss (1.20), and widening to 32 slightly overfits --- that's our primary reason for
16. But the more revealing result is that out-of-sample \textbf{regime coverage
does not improve with capacity}: the validation period populates only two of the
four regimes at \emph{every} width. So the out-of-sample collapse is a genuine
structural shift, not too small a bottleneck --- which is exactly why the trading
rule struggles later.
\end{cue}

\begin{cue}{Clustering into regimes}{40 sec}
Finally, we cluster. I extract the 16-D latent vector for every window and apply
\textbf{K-Means with K=4}, fitting the centres on \emph{training} latents only and
assigning val/test out of sample --- same anti-leakage logic as the scaler. Four
clusters is a deliberate choice: the report is honest that cluster-validity metrics
don't single out K=4, so we pick it on \textbf{economic} grounds --- four states map
cleanly onto easing, inflationary, stress, and a transitional baseline.
\end{cue}

\begin{cue}{Hand over}{20 sec}
So we have a non-linear 16-D representation and four regimes. The natural question
is: are they any better than a simple linear baseline, and do they mean anything?
That's Adene.
\end{cue}

% =====================================================================
% PERSON 3
% =====================================================================
\setp{p3}{p3l}
\section{Person 3 — PCA, Trading Test \& Outlook (Report §V-C–F; §VI–VIII)}
\personbanner{Adene Dinoshi}{PCA Comparison, Strategy, Limitations \& Future Work}{$\approx$ 5 min}

\begin{numbers}{Key numbers (from the report)}
\sffamily\small
\textbf{16-component} PCA $\to$ \textbf{70.8\%} train variance (10 PC = 67\%, 50 PC = 82\%) \;$\cdot$\;
Agreement \textbf{ARI = 0.52}, \textbf{NMI = 0.58} \;$\cdot$\;
Crisis regime $\approx$ \textbf{12\%} of windows \;$\cdot$\;
Trading Sharpe \textbf{train 0.23 / val 0.13 / test $-$0.28} \;$\cdot$\;
Signal-robustness check: \textbf{every} feature set loses out of sample.
\end{numbers}

\begin{cue}{Opening}{20 sec}
Thank you, Eleni. I'm Adene. My part is the payoff of the report: how the
autoencoder compares to a PCA baseline (Section~V-C,D), whether the regimes carry
\emph{economic} value in a trading test (Section~V-E), how robust that result is,
and where the project's limits and next steps lie.
\end{cue}

\begin{slide}\reportref{Fig.~3 ``01\_scree\_plot'' --- PCA explained variance}\end{slide}
\begin{cue}{The PCA baseline}{40 sec}
PCA is the natural \textbf{linear benchmark}: fast, interpretable, well understood.
If the autoencoder can't beat it, the extra complexity isn't justified. To keep the
comparison fair I use the \emph{same} 16-dimensional projection, fit on training
only, and the same K-Means. The scree plot, Figure~3, shows 16 components retain
\textbf{70.8\%} of the variance --- so both methods work from a genuinely compressed
view, and any difference reflects \emph{linear vs.\ non-linear}, not dimensionality.
\end{cue}

\begin{slide}\reportref{Fig.~4 ``07\_regime\_agreement\_heatmap''}\end{slide}
\begin{cue}{Agreement --- the key methodological result}{50 sec}
How much do the two methods agree? Section~V-C reports an \textbf{Adjusted Rand
Index of 0.52} and \textbf{Normalised Mutual Information of 0.58} --- \emph{moderate}
agreement, shown in the heatmap (Figure~4). This is our central finding: PCA and
the autoencoder share real regime structure but disagree on roughly half the
boundaries. The honest reading, which the report stresses, cuts both ways: the
autoencoder \emph{does} capture non-linear structure PCA can't reproduce in 16
components --- but a large share of the regime signal is in fact linear, so the
neural network's added value here is real yet modest.
\end{cue}

\begin{slide}\reportref{Fig.~5 ``05\_regime\_timeline'' + Table~IV (regime profiles)}\end{slide}
\begin{cue}{What the regimes mean}{45 sec}
Section~V-D interprets the four regimes from their average feature values.
\textbf{Regime~0}: low short and long rates --- a calm, easing environment.
\textbf{Regime~2}: high CPI and rising 10-year yields --- inflationary pressure.
\textbf{Regime~3} is the most distinct --- peak yield-spread stress and unemployment
with the sharpest GDP contraction: our \textbf{crisis} regime, about 12\% of
windows. \textbf{Regime~1} is the large, diffuse baseline. The timeline (Figure~5)
backs this up: Regime~3 clusters around the early-90s recession, 2008, and the COVID
shock.
\end{cue}

\begin{slide}\reportref{Table~V (position rule) + Table~VI (strategy Sharpe by split)}\end{slide}
\begin{cue}{Do the regimes carry economic value?}{55 sec}
This is the question that ties back to ``Investments''. In Section~V-E we build a
deliberately simple test: rank the regimes by their average forward return
\emph{in training}, then go \textbf{long} the best regime, \textbf{short} the worst,
flat otherwise --- and \textbf{freeze that rule} and apply it unchanged out of
sample. In sample it works as designed: Sharpe \textbf{0.23}, and the crisis regime
correctly precedes the worst returns. Out of sample the story falls apart:
validation Sharpe is a thin \textbf{0.13}, and on the test set the strategy
\emph{loses} money --- Sharpe \textbf{$-0.28$}, beaten by simple buy-and-hold, with
an information coefficient indistinguishable from zero. So the honest answer is:
\textbf{not yet a tradeable signal} on its own.
\end{cue}

\begin{slide}\reportref{Table~VII ``robustness to the choice of input signals''}\end{slide}
\begin{cue}{Is that failure robust?}{30 sec}
We checked whether this is just an artifact of our feature choice. Section~V-F
retrains the whole pipeline on three input sets --- macro-only, risk/sentiment-only,
and the full 30 features. The verdict is unchanged: \emph{every} set is profitable
in sample but posts a \textbf{negative test Sharpe}. The out-of-sample breakdown is
a property of the regime signal over this period, not of one feature selection.
\end{cue}

\begin{cue}{Limitations \& future work}{45 sec}
The report is upfront about limits (Sections~VI–VII). The windows overlap heavily
--- consecutive inputs share 25 of 26 weeks --- so the effective sample is far
smaller than 1365 and the regimes look more stable than they are. The autoencoder is
trained only to \emph{reconstruct}, so the regimes are a by-product of compression,
not optimised to predict returns. And the Sharpe figures are on standardised, not
raw percentage, returns. The natural next steps follow directly: a
clustering-aware or contrastive training objective, LSTM/variational architectures,
combining the regime with confirming predictors instead of trading it alone, and
\textbf{periodic retraining} so the regimes adapt instead of being frozen at 2014.
\end{cue}

\begin{cue}{Conclusion}{30 sec}
To conclude: Harjot built a leak-free pipeline over 38 years of weekly data; Eleni
trained a non-linear autoencoder compressing it to 16 dimensions and clustered four
regimes; and the PCA benchmark shows the autoencoder adds genuine but modest
non-linear value. The regimes are a useful \textbf{descriptive} lens on market
history --- but a frozen regime rule does not survive the post-2020 structural
break. Representation learning characterises regimes well; turning that into an
investable signal needs adaptation, not just a wider bottleneck. Thank you ---
we're happy to take questions.
\end{cue}

\newpage
% #####################################################################
% PART II — Q&A
% #####################################################################
\setp{ink}{soft}
\begin{center}
{\sffamily\bfseries\Huge\color{ink} Part II \;\textemdash\; Q\&A Preparation}
\end{center}
\vspace{2pt}
\begin{tcolorbox}[enhanced, sharp corners, colback=amberl, colframe=amber,
    boxrule=0pt, leftrule=4pt, left=10pt, right=10pt, top=6pt, bottom=6pt]
\sffamily\small \textbf{\textcolor{amber}{Tip.}} Every answer below maps to a claim
in the report. Know your own section cold; skim the others so you can field a
hand-off question. If unsure, it's fine to say ``that's more Eleni's area'' and
redirect.
\end{tcolorbox}

% --------------------- Person 1 Q&A ----------------------------------
\setp{p1}{p1l}
\section{Q\&A — Person 1 (Data, §III)}

\begin{qa}{Why weekly frequency, not monthly or daily?}
Weekly has enough resolution to catch transitions as they develop --- monthly might
already be mid-transition --- while avoiding daily microstructure noise. Macro series
like CPI and GDP are monthly/quarterly, so they're forward-filled to weekly, which is
fine for slow-moving signals.
\end{qa}

\begin{qa}{Why log returns instead of simple percentage returns?}
Log returns are additive over time, approximately normal for small moves, and the
standard in financial econometrics. The numerical difference is tiny at this scale,
but it makes the ADF stationarity tests cleaner --- the eight price columns go from
$p\approx 1.0$ to $p<0.0001$ --- and is methodologically correct.
\end{qa}

\begin{qa}{You split 70/15/15 by time --- why not k-fold cross-validation?}
K-fold shuffles time, so the model would train on the future and validate on the
past --- forward-looking bias that invalidates any out-of-sample claim. Financial
data is time-ordered; forward chaining (train oldest → test newest) honestly
simulates production, and it's what makes our test-set Sharpe meaningful.
\end{qa}

\begin{qa}{Why a 26-week window? Would a different size change the regimes?}
A regime is a stretch of time, not a single week, so we need a window. We picked 26
weeks ($\approx$ 6 months) on economic grounds --- a natural half-year horizon ---
long enough to smooth single-week noise but short enough to catch a change within a
quarter or two, rather than over-fitting the window length to the data.
\end{qa}

\begin{qa}{How exactly does your split prevent leakage?}
The \texttt{StandardScaler} is fit on \emph{training data only} and then applied to
validation and test. If we had fit it on all data, every split would have mean
$\approx 0$; instead validation and test means \emph{drift} away from zero, which is
the visible proof the scaler never saw the future. K-Means and PCA are likewise fit
on train only.
\end{qa}

\begin{qa}{Which 8 columns become log returns, and why those?}
The eight price/index \emph{levels} --- equity market, gold, the dollar, corporate
bonds, Treasury, oil, and the value/growth series. They trend, so they're
non-stationary as levels. The other 22 are already stationary rates/ratios (yields,
CPI, CAPE, unemployment\ldots) so we keep them as-is.
\end{qa}

\begin{qa}{Is 38 years of data enough?}
It spans multiple business cycles, two major bear markets (2000–02, 2008–09), two
inflation episodes, and several credit-stress events --- reasonable for studying
multi-year macro regimes. It is \emph{not} long enough for strong claims about rare
tail events, which the report explicitly does not make.
\end{qa}

% --------------------- Person 2 Q&A ----------------------------------
\setp{p2}{p2l}
\section{Q\&A — Person 2 (Autoencoder, §IV–V-B)}

\begin{qa}{Why a fully-connected autoencoder, not an LSTM or Transformer?}
The input is a fixed-length 26-week window flattened to 780 values, and we want the
overall \emph{state} of that window, not a next-step prediction, so order inside the
window matters less. A dense AE is a natural, efficient baseline; LSTM and
Transformer variants are listed in the report's future work.
\end{qa}

\begin{qa}{Why 16 latent dimensions --- and how do you justify it?}
Two reasons in the report. First, the features are highly correlated and a
16-component PCA already explains $\approx$ 71\% of variance, so effective
dimensionality is well below 30. Second, the ablation (Table~III) shows 16 gives the
lowest \emph{test} reconstruction loss (1.20); going to 32 slightly overfits.
Matching the PCA dimension also keeps the comparison fair.
\end{qa}

\begin{qa}{What is early stopping and why does it matter here?}
It monitors validation loss each epoch and keeps the best checkpoint, stopping after
40 epochs with no improvement. Without it the model keeps memorising the training
set --- training loss falls while validation loss rises. Since we want regime
structure that generalises across time, we select the model best on held-out data.
\end{qa}

\begin{qa}{Why is the test reconstruction loss (1.20) so much higher than train (0.30)?}
The test period (2020–2026) includes COVID, the fastest inflation in 40 years, and
an aggressive Fed hiking cycle --- combinations the 1988–2014 training era never saw
together. The high reconstruction loss means those windows are genuinely novel. We
read it as a regime shift, not a failure --- and the ablation confirms it persists at
every latent width.
\end{qa}

\begin{qa}{Why K-Means with K=4? Did you try other K?}
Honestly, the cluster-validity metrics don't single out K=4 --- silhouette rises
monotonically from about 0.25 at K=2 to 0.35 at K=8, and the inertia elbow is
shallow. So K=4 is chosen on \emph{economic} grounds: four states map cleanly to
easing, inflationary, stress, and a transitional baseline. We trade a little
silhouette for interpretability and out-of-sample stability, because larger K
fragments into clusters that are harder to interpret and less stable OOS.
\end{qa}

\begin{qa}{Why fit K-Means only on training, then predict val/test?}
Same logic as the scaler: we can't use the future to \emph{define} regimes.
Clustering on all data would let val/test points move the centroids, leaking future
information into the labels. Fitting on training only makes val/test assignments true
out-of-sample predictions.
\end{qa}

\begin{qa}{The validation period only uses two of the four regimes --- isn't that a bug?}
No --- and the ablation (Table~III) is the key evidence. The validation window
collapses to two regimes whether the bottleneck is 2- or 32-dimensional, so it's not
insufficient capacity; it's a genuine structural shift. The out-of-sample periods
simply don't exercise the full regime vocabulary learned in training.
\end{qa}

% --------------------- Person 3 Q&A ----------------------------------
\setp{p3}{p3l}
\section{Q\&A — Person 3 (PCA, Strategy \& Limits, §V-C–VIII)}

\begin{qa}{What does ``ARI = 0.52'' actually mean?}
The Adjusted Rand Index measures how often both methods agree on whether two windows
are in the same regime, corrected for chance: 1 = perfect, 0 = random. At 0.52 they
share real structure but disagree on about half the boundaries --- where they agree
the signal is the same; where they differ, the AE's non-linearity groups things
differently than linear PCA.
\end{qa}

\begin{qa}{If much of the structure is linear, was the autoencoder worth it?}
That's exactly the honest tension the report names. The AE adds something --- ARI is
well below 1, so it finds boundaries PCA can't --- but with 71\% of variance already
linear, the gain is \emph{modest}. The contribution is the rigorous, fair comparison
that quantifies how much non-linearity actually buys, rather than assuming a neural
net is better.
\end{qa}

\begin{qa}{How do you know the regime labels map to real economic states?}
From the mean standardised feature values per regime (Table~IV) plus the timeline. A
regime with high CPI and rising 10-year yields, whose windows cluster around known
inflationary periods, is a consistent story; the crisis regime overlaps known
recession dates. These are reasoned inferences, not ground truth --- none exists for
unlabelled regimes.
\end{qa}

\begin{qa}{Walk me through the trading strategy --- is this a real backtest?}
It's a deliberately simple \emph{diagnostic}, not a deployable strategy. We rank the
four regimes by average forward return \emph{in training}, go long the best and short
the worst, freeze that rule, and apply it unchanged to val/test. In sample Sharpe is
0.23; out of sample it's 0.13 then $-0.28$ on test, losing to buy-and-hold. There's
no position sizing, transaction costs, or risk management --- it only asks whether the
regimes carry \emph{directional} information.
\end{qa}

\begin{qa}{Why does the strategy work in sample but fail out of sample?}
By construction it's profitable in sample --- the rule is fit to the training
ranking. The out-of-sample failure lines up with everything else: the high test
reconstruction loss and the regime collapse all point to a post-2020 structural
break. A rule frozen on 1988–2014 simply doesn't describe 2020–2026.
\end{qa}

\begin{qa}{Couldn't the failure just be your particular 30 features?}
We tested that (Table~VII). Retraining on macro-only, on risk/sentiment-only, and on
the full set, \emph{all three} are profitable in sample yet post a negative test
Sharpe. So the breakdown is robust to the input choice --- it's the regime signal
over this period, not the feature set.
\end{qa}

\begin{qa}{Your Sharpe ratios look small --- are they comparable to real strategies?}
No, and the report flags this. Sharpe and IC are computed on \emph{standardised}
log-returns, not raw percentage returns, so the magnitudes aren't directly comparable
to a real percentage-return strategy. The sign- and ranking-based conclusions (in vs.\
out of sample) are unaffected; the absolute numbers are not the point.
\end{qa}

\begin{qa}{What's the single biggest limitation?}
The overlapping windows. With a 26-week window and a one-week stride, consecutive
inputs share 25 of 26 weeks, so the 1365 training windows correspond to far fewer
\emph{independent} episodes. Effective sample size is much smaller than the row
count, which inflates apparent regime stability and makes any regime–return
relationship prone to overfitting.
\end{qa}

\begin{qa}{If you had more time, what's the first thing you'd change?}
Add a \textbf{clustering-aware or contrastive term} to the training objective. Right
now the AE only minimises reconstruction error, so the regimes are a by-product of
compression rather than optimised for separation or for predicting returns. Closely
behind: periodic retraining so the regimes adapt instead of being frozen at 2014.
\end{qa}

% --------------------- Cross-team Q&A --------------------------------
\setp{ink}{soft}
\section{Q\&A — Cross-Team Questions}
\setp{amber}{amberl}

\begin{qa}{So what is the one-sentence takeaway of the whole report?}
Representation learning gives a useful \emph{descriptive} lens on historical market
regimes --- the autoencoder adds real but modest non-linear value over PCA --- but a
regime rule frozen on history does not survive the post-2020 structural break, so it
is not yet a stand-alone investment signal.
\end{qa}

\begin{qa}{How does any of this connect to real investment decisions?}
That's precisely why we added the trading test in Section~V-E: it converts an
abstract clustering result into an economic question --- do the regimes precede
different returns? In sample yes, out of sample no. The honest conclusion is that
this is a research tool, and a path to an investable version (confirming predictors,
adaptation) is laid out in future work.
\end{qa}

\begin{qa}{Why not Hidden Markov Models, the classic regime method?}
HMMs assume a Markov transition structure and need the number of states and emission
distributions specified up front. Autoencoders and PCA learn the representation
purely from data --- which is exactly the linear-vs-non-linear comparison this
project set out to make. HMMs are a natural extension.
\end{qa}

\begin{qa}{If the strategy loses out of sample, why present it at all?}
Because a negative result, honestly reported, is the finding. It shows the regimes
are economically \emph{sensible} in sample (the crisis regime precedes the worst
returns) and pinpoints \emph{where} generalisation breaks --- the post-2020 regime
shift --- which is far more useful than a cherry-picked in-sample Sharpe.
\end{qa}

\begin{qa}{Is the whole pipeline reproducible?}
Yes --- fixed random seed, train-only fitting at every stage, and a documented
data handoff, so the figures and tables in the report regenerate end to end. The
trading and signal-robustness results come from scripts in the
\texttt{trading-strategy/} folder that read the same frozen windows and regime
labels.
\end{qa}

\vspace{10pt}
\begin{center}
{\color{rule}\rule{0.6\textwidth}{1pt}}\\[6pt]
{\sffamily\small\color{ink} End of document \;$\cdot$\; good luck with the presentation!}
\end{center}

\end{document}
